Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Pushing on related work.

\item {\bf maze}: generates a random maze and finds a path (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
  Makes use of Scheme's {\tt call-with-current-continuation}
  operation.

\item {\bf church}: tests distributivity of multiplication over
  addition for a few Church numerals (introduced by
  \cite{dvanhorn:Vardoulakis2011CFA2}).

\item {\bf lattice}: enumerates the maps between two lattices (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

\item {\bf boyer}: a term-rewriting theorem prover (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It makes extensive use of higher-order procedures and is
  written in a continuation-passing style (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})

\end{enumerate}

@@

[Computes results more like Earl, et al.'s 0-CFA, which times out on
  the Church numeral benchmark and so even though it offers a fair
  point of comparison, a more thorough evaluation is likely to
  uninformative as the other benchmarks are highly likely to timeout
  as well (and it would require significant effort to extend their
  implementation with the features needed to analyze our benchmark
  suite).  That implementation is evaluated against much smaller
  benchmarks: the largest program consists of 63 \emph{subexpressions}
  and times out after 30 minutes.]

@@

[Should probably talk about mCFA too.]

@@

This work clearly closely follows Van Horn and Might's original papers
on abstracting abstract
machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic},
which in turn is one piece of the large body of research on flow
analysis for higher-order languages (see
Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough
survey).  The AAM approach sits at the confluence of two major lines
of research: (1) the study of abstract
machines~\cite{dvanhorn:landin-64} and their systematic
construction~\cite{dvanhorn:reynolds-hosc98}, and (2) the theory of
abstract interpretation
\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot:1979:Galois}.

@@

This effort increases substantially as the complexity of the analyzed
language increases.  Both the work of maintaining the concrete
semantics and constraint system (and the relations between them) must
be scaled simultaneously.  However, constraint systems, which have
been extensively studied in their own right, enjoy efficient
implementation techniques and can be expressed in declarative logic
languages that are heavily
optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
Consequently, constraint-based analyses can be computed quickly.  For
example, Jagannathan and Wright's polymorphic splitting
implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
the church numeral example of figure~\ref{fig:church} in 7
milliseconds, about 25 times faster than the fastest implementation
considered here.  These analyses compute very different things (as
detailed in section~\ref{sec:accept}), so the performance comparison
is not apples-to-apples.  We view the primary contribution of this
work as a systematic path that eases the design, verification, and
implementation of analyses using the abstracting abstract machine
approach to within a factor of performant constraint-based analyses.

>>>>

\item {\bf maze}: generates a random maze using Scheme's {\tt
  call/cc} operation and finds a path solving
  the maze (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf church}: tests distributivity of multiplication over
  addition for Church numerals (introduced by
  \cite{dvanhorn:Vardoulakis2011CFA2}).

\item {\bf lattice}: enumerates the maps between two lattices (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf boyer}: a term-rewriting theorem prover (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It makes extensive use of higher-order procedures and is
  written in a continuation-passing style (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

The analysis considered here computes results more like Earl, et al.'s
0-CFA, which times out on the \Church{} benchmark because it does not
widen the store as described for our baseline evaluator.  So even
though it offers a fair point of comparison, a more thorough
evaluation is probably uninformative as the other benchmarks are
likely to timeout as well (and it would require significant effort to
extend their implementation with the features needed to analyze our
benchmark suite).  That implementation is evaluated against much
smaller benchmarks: the largest program is 30 lines.

@@

@@

This work clearly closely follows Van Horn and Might's original papers
on abstracting abstract
machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic},
which in turn is one piece of the large body of research on flow
analysis for higher-order languages (see
Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough
survey).  The AAM approach sits at the confluence of two major lines
of research: (1) the study of abstract
machines~\cite{dvanhorn:landin-64} and their systematic
construction~\cite{dvanhorn:reynolds-hosc98}, and (2) the theory of
abstract interpretation
\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot1979Systematic}.

@@

This effort increases substantially as the complexity of the analyzed
language increases.  Both the work of maintaining the concrete
semantics and constraint system (and the relations between them) must
be scaled simultaneously.  However, constraint systems, which have
been extensively studied in their own right, enjoy efficient
implementation techniques and can be expressed in declarative logic
languages that are heavily
optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
Consequently, constraint-based analyses can be computed quickly.  For
example, Jagannathan and Wright's polymorphic splitting
implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
the \Church{} benchmark about 25 times faster than the fastest
implementation considered here.  These analyses compute very different
things, so the performance comparison is not apples-to-apples.

The AAM approach, and the state transition graphs it generates, encode
temporal properties not found in classical constraint-based analyses
for higher-order programs.
%
These analyses (ultimately) compute judgments on program terms and
contexts, e.g., at expression $e$, variable $x$ may have value $v$.
%
The judgments do not relate the order in which expressions and context
may be evaluated in a program, e.g., it has nothing
to say with regard to question like, ``Do we always evaluate $e_1$
before $e_2$?'' or ``Is it always the case that a file handle is
opened, read and then closed in that order?''
%
The state transition graphs can answer these kinds of queries, but
this does not come for free: respecting temporal order imposes an
order in which states and terms may be evaluated \emph{during} the
analysis.

We view the primary contribution of this work as a systematic path
that eases the design, verification, and implementation of analyses
using the abstracting abstract machine approach to within a factor of
performant constraint-based analyses.
